---
title: "Microbiota-based SCReening of Anal Cancer in HIV-infected individuals (SCRAtCH) - discovery_vs_validation"
author: "Sajan Raju"
date: "3/2/2023"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

# Load R packages

<br><br>
<h4> Below mentioned R packages are used in this analyses</h4>

```{r loadPackages, echo=TRUE, warning=FALSE, message=FALSE}
library(phyloseq)
library(qiime2R)
library(ggplot2)
library(knitr)
library(microbiome)
library(table1)
library(ggpubr)
library(hrbrthemes)
library(vegan)
library(microbiomeutilities)
#library(microbiomeMarker)
library(tidyverse)
library(ggstatsplot)
library(haven)
library(ggsci)
library(patchwork)
installed.packages()[names(sessionInfo()$otherPkgs), "Version"]

```
<br><br>

# Import QIIME files and metadata in R

```{r importingInput, echo=TRUE, warning=FALSE, message=FALSE}
setwd("~/OneDrive - Universitetet i Oslo/scratch_data/qiime_analysis/")
physeq <- qza_to_phyloseq(
  features="~/OneDrive - Universitetet i Oslo/scratch_data/qiime_analysis/table-all-dada2.qza",
  tree="~/OneDrive - Universitetet i Oslo/scratch_data/qiime_analysis/rooted-tree.qza",
  taxonomy="~/OneDrive - Universitetet i Oslo/scratch_data/qiime_analysis/taxonomyV3V4.qza",
  metadata = "~/OneDrive - Universitetet i Oslo/scratch_data/qiime_analysis/hHSILnew.tsv" )

physeq
```

# ASV/OTU details

```{r ASVdetail, echo=TRUE, warning=FALSE, message=FALSE}
# physeq

# sample_data(physeq)$cohort <- factor(sample_data(physeq)$cohort, levels = c("No_HSIL", "HSIL"))
# sample_data(physeq)$bAINcat <- factor(sample_data(physeq)$bAINcat, levels = c("Normal", "LSIL","HSIL"))
# physeq
#saveRDS(physeq, "scratchphylo.Rdata")
#readRDS("scratchphylo.Rdata")
#save(physeq, file = "phylodata.RData")
#load("data.RData")
#Color palette to use Set2
#(LSIL-Green) "#66C2A5" 
#(HSIL-Red) "#FC8D62" 
#(Normal- Blue) "#8DA0CB"

```

# Subset data for validation cohort

<br> Removed unwanted taxa to keep only bacteria.

```{r subsetting, echo=TRUE, message=FALSE,warning=FALSE,}
#physeq_bact<-subset_samples(physeq, deriv=="v")
physeq_bact <- physeq
#physeq_bact<-subset_samples(physeq_bact, cohort=="HSIL"  | cohort=="No_HSIL")
#physeq_bact
#physeq_bact<-subset_samples(physeq_bact, bAINcat=="HSIL"  | bAINcat=="LSIL" | bAINcat=="Normal")
physeq_bact
physeq_bact = subset_taxa(physeq_bact, Kingdom=="d__Bacteria")
physeq_bact

pseq<-physeq_bact
ps1 <- prune_taxa(taxa_sums(pseq) > 0, pseq)
ps1
```

<h4>Table showing the distribution of cohort groups in samples </h4> 

```{r table1, echo=TRUE, message=FALSE,warning=FALSE,}
newdata2<-meta(ps1)
table1::table1(~ cohort, data = newdata2)

```

<br><br>

# Alpha diversity - (simple box plot)

<p> A comprehensive list of global indicators of the diversity showed below in  the figure, which includes various measures of richness, evenness, diversity, dominance, and rarity.
<br>
```{r alphall, echo=TRUE, message=FALSE,warning=FALSE,fig.height=5, fig.width=10}
#prune_samples(physeq_bact, "discovery")
alpha_meas = c("Observed", "Chao1", "ACE", "Shannon", "InvSimpson")
p <- plot_richness(ps1, "cohort", measures=alpha_meas)
p + geom_boxplot(data=p$data, aes(x=cohort, y=value, color=NULL), alpha=0.1) +
    ggtitle("Diversity measures in discovery and validation cohorts ") +
    stat_compare_means(method = "wilcox.test") 
# + geom_boxplot(width = .2, outlier.shape = NA, coef = 0)
tab <- microbiome::alpha(ps1, index = "all")

```
<br><br>

# Alpha diversity - ggstatplot  in two cohorts

```{r ggstat,echo=TRUE, message=FALSE,warning=FALSE,fig.height=5, fig.width=10}
################## alpha diversity

tab2<-estimate_richness(ps1, measures=c("Observed", "ACE", "InvSimpson", "Shannon", "Chao1", "Fisher"))
ps1.meta <- meta(ps1)
ps1.meta$Observed <- tab2$Observed 
ps1.meta$Chao1 <- tab2$Chao1
ps1.meta$ACE <- tab2$ACE
ps1.meta$Fisher <- tab2$Fisher

ps1.meta$Shannon <- tab2$Shannon 
ps1.meta$InverseSimpson <- tab2$InvSimpson


pfish<-ggbetweenstats(ps1.meta, x=cohort, y=Fisher, type = "nonparametric",plot.type = "violin")+
  labs(
   x = "cohorts",
    y = "Fisher index",
      palette = "Set2")+
    #title = " Fisher index")  +
  theme(
    axis.ticks = element_blank(),
    axis.line = element_line(colour = "grey50"),
    panel.grid = element_line(color = "#b4aea9"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linetype = "dashed"),) 
# + scale_color_lancet()
#pfish

pshan<-ggbetweenstats(ps1.meta, x=cohort, y=Shannon, type = "nonparametric",plot.type = "violin")+
  labs(
   x = "cohorts",
    y = "Shannon index",
      palette = "Set2")+
   # title = " Shannon diversity index")  +
  theme(
    axis.ticks = element_blank(),
    axis.line = element_line(colour = "grey50"),
    panel.grid = element_line(color = "#b4aea9"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linetype = "dashed"),) 
# + scale_color_lancet()
#pshan

pobs<-ggbetweenstats(ps1.meta, x=cohort, y=Observed, type = "nonparametric",plot.type = "violin")+
  labs(
   x = "cohorts",
    y = "Observed richness",
      palette = "Set2")+
    #title = " Observed richnes index")  +
  theme(
    axis.ticks = element_blank(),
    axis.line = element_line(colour = "grey50"),
    panel.grid = element_line(color = "#b4aea9"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linetype = "dashed"),) 
# + scale_color_lancet()
#pobs

pchao<-ggbetweenstats(ps1.meta, x=cohort, y=Chao1, type = "nonparametric",plot.type = "violin")+
  labs(
   x = "cohorts",
    y = "Chao1 index ",
      palette = "Set2")+
    #title = " Chao1 index ")  +
  theme(
    axis.ticks = element_blank(),
    axis.line = element_line(colour = "grey50"),
    panel.grid = element_line(color = "#b4aea9"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linetype = "dashed"),) 
# + scale_color_lancet()
#pchao

pace<-ggbetweenstats(ps1.meta, x=cohort, y=ACE, type = "nonparametric",plot.type = "violin")+
  labs(
   x = "cohorts",
    y = "ACE index ",
      palette = "Set2")+
    #title = " ACE index ")  +
  theme(
    axis.ticks = element_blank(),
    axis.line = element_line(colour = "grey50"),
    panel.grid = element_line(color = "#b4aea9"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linetype = "dashed"),) 
# +   scale_color_lancet()
#pace

pinv<-ggbetweenstats(ps1.meta, x=cohort, y=InverseSimpson, type = "nonparametric",plot.type = "violin")+
  labs(
   x = "cohorts",
    y = "Inverse Simpson index ",
   palette = "Set2")+
   # title = " Inverse Simpson index")  +
  theme(
    axis.ticks = element_blank(),
    axis.line = element_line(colour = "grey50"),
    panel.grid = element_line(color = "#b4aea9"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linetype = "dashed"),) 
#+   scale_color_lancet()
#pinv


## combining the individual plots into a single plot
pcomb<-combine_plots(
  list(pobs, pchao, pshan,pinv, pfish),
  #list(pobs, pchao, pace, pshan,pinv, pfish),
  plotgrid.args = list(nrow = 2),
  annotation.args = list(
    title = "Diversity measures in discovery and validation cohorts ")
    #caption = "Source: Gapminder Foundation"
)
```


```{r ggstatplot,echo=TRUE, message=FALSE,warning=FALSE,fig.height=5, fig.width=10}
pcomb
ggsave("ggstat_alpha_allplots1_inCohorts.pdf", width = 42, height = 20, units = "cm", dpi = 600)
ggsave("ggstat_alpha_allplots1_inCohorts.jpg",bg="white",width=18, height=12, dpi = 500)

```




# Relative abundance - Family and Genus level 


```{r taxafg, fig.height=10, fig.width=10, echo=TRUE, fig.align="center", out.extra="id=logo"}
library(RColorBrewer)
# Define the number of colors you want
nb.cols <- 18
mycolors <- colorRampPalette(brewer.pal(8, "Paired"))(nb.cols)

ps2 <- ps1 %>% aggregate_top_taxa2(level = "Family", top = 10) %>% microbiome::transform(transform = "compositional")
pseq2<-ps2
p4 <- plot_composition(pseq2,  taxonomic.level = "Family", sample.sort = "cohort", group_by="cohort",
                       x.label = "cohort") + guides(fill = guide_legend(ncol = 1)) +  
  scale_fill_manual(values = mycolors) +
  #scale_fill_brewer("Family", palette = "Paired") +
  scale_y_percent() + labs(x = "Samples", y = "Relative abundance (%)",
                           title = "Relative abundance of family in two cohorts")+ theme_ipsum(grid="Y")+ theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

print(p4)

ggsave("relabund_family_all_inCohorts.jpg",bg="white",width=10, height=8)
#ggsave("relabund_family_all_inCohorts.pdf", width = 42, height = 20, units = "cm", dpi = 600)
#ggsave("relabund_family_all_inCohorts.pdf")
# Averaged by group
p44 <- plot_composition(pseq2, average_by = "cohort", transform = "compositional") +
#  scale_fill_brewer("Family", palette = "Paired") + 
    scale_fill_manual(values = mycolors) +
theme_ipsum(grid="Y") +
  theme(axis.text.x = element_text(angle=0, hjust=1), legend.text = element_text(face = "italic"))

print(p44)

ggsave("relabund_family_combined_inCohorts.jpg",bg="white",width=8, height=8)
#ggsave("relabund_family_combined_inCohorts.pdf", width = 42, height = 20, units = "cm", dpi = 600)
#ggsave("relabund_family_combined_inCohorts.pdf")
##### GENUS

ps3 <- ps1 %>% aggregate_top_taxa2(level = "Genus", top = 15) %>% microbiome::transform(transform = "compositional")
pseq2<-ps3
p5 <- plot_composition(pseq2,  taxonomic.level = "Genus", sample.sort = "cohort", group_by="cohort",
                       x.label = "cohort") + guides(fill = guide_legend(ncol = 1)) +   
  #scale_fill_brewer("Genus", palette = "Paired") +
  scale_fill_manual(values = mycolors) +
  scale_y_percent() + labs(x = "Samples", y = "Relative abundance (%)",
                           title = "Relative abundance of genus in two cohorts")+ 
  theme_ipsum(grid="Y")+ theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

print(p5)

ggsave("relabund_Genus_all_inCohorts.jpg",bg="white",width=10, height=8)
#ggsave("relabund_Genus_all_inCohorts.pdf", width = 42, height = 20, units = "cm", dpi = 600)
#ggsave("relabund_Genus_all_inCohorts.pdf")

# Averaged by group
p55 <- plot_composition(pseq2, average_by = "cohort", transform = "compositional") +
#  scale_fill_brewer("Genus", palette = "Paired") + 
    scale_fill_manual(values = mycolors) +
theme_ipsum(grid="Y") +
  theme(axis.text.x = element_text(angle=0, hjust=1), legend.text = element_text(face = "italic"))

print(p55)

ggsave("relabund_Genus_combined_inCohorts.jpg",bg="white",width=8, height=8)
#ggsave("relabund_Genus_combined_inCohorts.pdf", width = 42, height = 20, units = "cm", dpi = 600)
#ggsave("relabund_Genus_combined_inCohorts.pdf")
```
# Relative abundance - phyla level 

```{r taxaphy, fig.height=10, fig.width=10, echo=TRUE, fig.align="center", out.extra="id=logo"}
transform <- microbiome::transform
# Merge rare taxa to speed up examples
pseq <- transform(ps1, "compositional")
pseq <- aggregate_rare(pseq, level = "Phylum", detection = 0.1/100, prevalence = 20/100)
pseq2<-pseq
p4 <- plot_composition(pseq2,
                       taxonomic.level = "Phylum",
                       sample.sort = "cohort", group_by="cohort",
                       x.label = "cohort") +
  guides(fill = guide_legend(ncol = 1)) +   scale_fill_brewer("Phylum", palette = "Paired") +
  scale_y_percent() +
  labs(x = "Samples", y = "Relative abundance (%)",
       title = "Relative abundance of taxonomic phylum")+
       #subtitle = "in validation cohort",
       #caption = "two cohorts.") + 
  theme_ipsum(grid="Y")+ 
  #theme(axis.text.x = element_text(angle=90, hjust=1))+
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())
print(p4)

# Averaged by group
p44 <- plot_composition(pseq2,
                      average_by = "cohort", 
                      transform = "compositional") +
  scale_fill_brewer("Phylum", palette = "Paired") +
  theme_ipsum(grid="Y") +
  theme(axis.text.x = element_text(angle=0, hjust=1),
        legend.text = element_text(face = "italic"))
print(p44)
#ggsave("relabund_PHYLA_combined_inCohorts.pdf", width = 42, height = 20, units = "cm", dpi = 600)


```


# Betadiversity in two cohorts - unweighted unifrac

<br>PERMANOVA quantifies multivariate community-level differences between groups.
<br>Permanova analysis used on unifrac distances and showed it in a NMDS plot. <br>
```{r betahsil, echo=TRUE, message=FALSE, warning=FALSE}
ordu.unwt.uni <- ordinate(ps1, "NMDS", "unifrac", weighted=F)

# check for Eigen values
# barplot(ordu.unwt.uni$values$Eigenvalues[1:10])

unwt.unifrac <- plot_ordination(ps1, ordu.unwt.uni, color="cohort")+ 
                                stat_ellipse(aes(group = cohort), linetype = 2)
unwt.unifrac <- unwt.unifrac + ggtitle("Unweighted UniFrac") + geom_point(size = 2.5)
            
unwt.unifrac <- unwt.unifrac + theme_classic() + scale_color_brewer("cohort", palette = "Set2")
print(unwt.unifrac)
ggsave("betadiv_nmds_unwtUnifrac_inCohorts.jpg",bg="white")
ggsave("betadiv_nmds_unwtUnifrac_inCohorts.pdf",dpi = 300)


metadf <- data.frame(sample_data(ps1))

unifrac.dist <- UniFrac(ps1,
                        weighted = TRUE,
                        normalized = TRUE,
                        parallel = FALSE,
                        fast = TRUE)
#Permanova analysis using Vegan (adonis)
permanova <- adonis2(unifrac.dist ~ cohort, data = metadf)

permanova
```

## betadispersion

Vegan::betadisper: Multivariate homogeneity of groups dispersions (variances)

```{r betadisphsil, echo=TRUE, message=FALSE, warning=FALSE}
beta2 <- betadisper(unifrac.dist, metadf$cohort)
permutest(beta2)

plot(beta2, main = "Ordination Centroids and Dispersion Labeled: Unweighted Unifrac Distance", sub = "")  

jpeg(file="centroiddispercohort_inCohorts.jpeg")
plot(beta2, main = "Ordination Centroids and Dispersion Labeled: Unweighted Unifrac Distance", sub = "")  
dev.off()

boxplot(beta2, main = "", xlab = "")

jpeg(file="centroidboxplotcohort_inCohorts.jpeg")
#pdf(file="centroidboxplotcohort_inCohorts.pdf")
boxplot(beta2, main = "", xlab = "")
dev.off()



```



# Differntial abundance - ASV level- DESeq2

```{r deseqphy, echo=TRUE, message=FALSE, error = FALSE,warning=FALSE}

# library(DESeq2)
# library(ggrepel)
# diagdds = phyloseq_to_deseq2(ps1, ~ cohort)
# 
# diagdds$cohort <- relevel(diagdds$cohort, ref = "No_HSIL")
# 
# dds<-estimateSizeFactors(diagdds, type = 'poscounts')
# #dds <- DESeq(dds, test="Wald")
# dds = DESeq(dds, test="Wald", fitType="parametric")
# 
# 
# res = results(dds, cooksCutoff = FALSE)
# alpha = 0.05
# #res
# sigtab1 = res[which(res$padj < alpha), ]
# sigtab1 = cbind(as(sigtab1, "data.frame"), as(tax_table(ps1)[rownames(sigtab1), ], "matrix"))
# head(sigtab1,20)
# 
# sigtab =as.data.frame(res)
# merged <- cbind(sigtab,tax_table(ps1))
# write.table(merged,"deseq_incohorts_table.tsv",sep = "\t")
# # add a column of NAs
# merged$diffexpressed <- "NO"
# # if log2Foldchange > 0.6 and pvalue < 0.05, set as "UP" 
# merged$diffexpressed[merged$log2FoldChange > 0.6 & merged$pvalue < 0.05] <- "UP"
# # if log2Foldchange < -0.6 and pvalue < 0.05, set as "DOWN"
# merged$diffexpressed[merged$log2FoldChange < -0.6 & merged$pvalue < 0.05] <- "DOWN"
# 
# pvolc<-ggplot( merged, aes(x=log2FoldChange, y= -log10(pvalue),col=diffexpressed)) + geom_point()+theme_minimal(base_size = 12, base_rect_size = 5)+ 
#   geom_vline(xintercept=c(-2.5, 2.5), col="black", linetype="dashed") +  
#   geom_hline(yintercept=-log10(0.05), col="blue", linetype= "dashed")+ scale_color_manual(values=c("dark green", "black", "dark red"))+ scale_y_continuous(breaks = seq(0, 45, by = 5))
# 
# pvolc
#   #geom_text_repel(data=merged, aes(label=Genus))
# ggsave("volcanopltdeseq_hsil_inCohorts.jpg",bg="white")

```

##  Differntial abundance analysis using LEFSE

```{r lefseFDR, echo=TRUE, message=FALSE, error = FALSE,warning=FALSE}
#   library(microbiomeMarker)
#   mm_lefse <- run_lefse(
#   ps1,  taxa_rank = "Genus",
#   wilcoxon_cutoff = 0.05,
#   group = "cohort",
#   kw_cutoff = 0.01,
#   multigrp_strat = TRUE,
#   strict=("1"),
#   lda_cutoff = 2
# )
# 
# mm_lefse
# marker_table(mm_lefse)
#plot_ef_bar(mm_lefse) + scale_color_brewer("cohort", palette = "Set2")
# #+ scale_fill_manual(values =c("below" = "#00bfc4", "above" = "#F8766D"))
```

# R session information
```{r}
sessionInfo()
```
